Cities must act to secure the future of urban cycling: our research shows how

### 2020-06-12 conversation.Rmd
Updated results - leeds + london
Add conversation draft

---
title: "Pop-up cycleways provide a peak into post COVID-19 transport futures"
---

As nations prepare to emerge from months of COVID-19 induced lockdowns, it is becoming clear life cannot simply return to normal. The pandemic has transformed all sectors of the economy, not least transport.

In the coming weeks and months, walking and cycling will play a vital role in enabling the reopening of many countries, as continued social distancing measures will mean that public transport options are likely to run at a fraction of their intended capacity. There is a window of opportunity here that could be missed if roads become busy and dangerous again before new active transport habits become embedded.

Local and national governments are scrambling to act. In the UK, £2 billion investment in measures to promote walking and cycling has been announced. £250 million of this is for "emergency interventions to make cycling and walking safer" as [Transport Secretary Grant Shapps put it](https://www.gov.uk/government/speeches/transport-secretarys-statement-on-coronavirus-covid-19-9-may-2020). New [guidance from the Department for Transport](https://www.gov.uk/government/publications/reallocating-road-space-in-response-to-covid-19-statutory-guidance-for-local-authorities/traffic-management-act-2004-network-management-in-response-to-covid-19) advocates "measures to reallocate road space to people walking and cycling".

@@

Local authorities urgently need answers on how this funding should be spent, and where road space reallocation should be prioritised. New cyclinginfrastructure which is developed based on rigorous analysis of the best available data, in combination with local knowledge and community engagement, are more likely to be effective than plans based on intuition alone. Recognising the importance of a strong and actionable evidence-base, and the window of opportunity, we used our experience developing tools that are already being used by local authorities, notably the multi-institution PCT and the CyIPT, to develop methods and outputs taylored for the new transport planning landscape.

<image id="334100" align="centre" caption="Screenshot from the website salfordliveablestreets.commonplace.is to support local responses to the COVID-19 pandemic, including the prioritisation of pop-up active transport infrastructure." />

In a pre-print [paper](https://osf.io/7wjb6) on methods to prioritise pop-up road space reallocation schemes, we have classified pop-up interventions into three categories. Firstly, in streets that are wide or have multiple lanes, roadspace can be reallocated to widen pavements or introduce pop-up cycleways, such as outside a hospital in [Leicester](https://www.leicestermercury.co.uk/news/leicester-news/light-lockdown-traffic-leicester-allows-4083625). Secondly, physical barriers can be used to make routes for walking and cycling [faster and more direct](https://cyclehighways.eu/design-and-build/infrastructure/filtered-permeability.html) than the routes for motor vehicles. This can be seen in Tower Hamlets, London, where simple barriers like planters are used to good effect. Thirdly, roads can be closed altogether to motor vehicles, as is happening in [Deansgate, Manchester](https://www.manchestereveningnews.co.uk/news/greater-manchester-news/part-deansgate-close-cars-city-18178175).

The first of these three options is the most simple to implement,  with relatively few implications for traffic flow. We have mapped out **link** the top locations for this reallocation of road space in ten major cities across England, with the intention of rolling out these methods across the country in the near future. Having identified roads where a spare lane could be re-purposed for cycling, we have assessed the cycle potential of these streets based on census data for travel to work and cycling uptake scenarios under the UK Government’s national target of doubling cycling levels by 2025. The length of road segments is also used to filter the top schemes.

The use of open data and publicly accessible models is a key aspect of this research. Given the global nature of the challenges we face, open and collaborative research is vital so that methods can be applied by as wide a user base as possible. The potential for international application can be seen for example in [research for the World Health Organisation](https://atfutures.github.io/upthat/articles/adaptation.html) modelling cycling uptake in cities in low income countries. Within the UK, [50+ Local Authorities](https://npct.github.io/pct-shiny/regions_www/www/static/03d_other_reports/2019-use-of-pct-report.pdf) are now using the PCT to inform their strategic cycle network plans.

Results from our [methods](https://osf.io/7wjb6) show that many cities have 'spare lanes' under the current levels of traffic in city centres and near key worker destinations such as hospitals. These results could support the prioritisation of vital resources dedicated to pop-up infrastructure. With the situation rapidly developing, some of the roads shown here, such as [Kirkstall Road](https://www.transportxtra.com/publications/local-transport-today/news/65382/leeds-to-pilot-enhanced-cycling-safety-u) (in the top 10 results our methods generated for Leeds), have already been identified by Local Authorities as targets for intervention. As the recovery continues, we hope our methods will support cities to take the opportunities that the crisis presents for active travel.

<image id="334113" align="centre" caption="Roads in Leeds with at least one spare lane that could be reallocate to provide space for active travel, and locations of hospitals" />

With the rush to act, there is a great need for evidence of where new interventions should be prioritised. As with the medical science, research is needed now. Methods developed to identify locations of high walking and cycling potential can help ensure that the ‘pop-up’ infrastructure that goes in is effective, safe, and placed where it is most needed. Agile research can respond to rapidly changing conditions, as government priorities shift and different sectors of the economy reopen. Measures can be undertaken on a temporary basis, but if action is prioritised in places where there is a long-term need for cycle improvements, there is a greater chance that they become permanent. In Paris, Covid-19 related measures aim to contribute towards Mayor Anne Hidalgo's long-term [Plan Velo](https://www.curbed.com/2020/1/15/21065343/bike-paris-cycling-anne-hidalgo).

@@

<!-- ![](https://raw.githubusercontent.com/cyipt/tempCycleways/master/README_files/figure-gfm/manchester-1.png) -->

>>>>

---
title: "Cities must act to secure the future of urban cycling: our research shows how"
# title: "Pop-up cycleways provide a peak into post COVID-19 transport futures"
output: github_document
---

Cities worldwide are preparing for the long transition out of [lockdown](https://www.bsg.ox.ac.uk/research/research-projects/coronavirus-government-response-tracker). Physical distancing measures will be in place for many months, with impacts on all walks of life, not least transport.

With public transport options running at low capacity and emerging evidence of the role of [air quality](https://www.sciencedirect.com/science/article/pii/S0048969720321215) and [exercise](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7118448/) in mitigating the risks of COVID-19, solutions are needed more than ever.

New cycleways are being introduced in [many cities](https://twitter.com/robinlovelace/status/1268834853472088066), allowing healthy habits started during the lockdown to continue. Transport authorities must act fast, however, to take advantage of the current [cycling boom](https://www.bbc.co.uk/news/business-52564351) while [traffic levels are still below normal](https://www.forbes.com/sites/carltonreid/2020/05/07/car-traffic-rise-of-11-shows-urgent-need-for-pop-ups-to-protect-cyclists-and-pedestrians/amp/), and to avoid [gridlock](https://static1.squarespace.com/static/5d30896202a18c0001b49180/t/5ee090e5635f3e0513867b40/1591775464939/Report_1+-+Congestion+ahead_+a+faster+route+is+now+available.pdf).

<image id="341146" align="centre" caption="New cycleway on space reclaimed from motor traffic as part of the COVID-19 response in Park Lane, London." source="Transport for London." />

Our research at the University of Leeds has identified roads where there is both space and demand for cycling infrastructure.

Our [methods](https://osf.io/7wjb6) have been used in a nationwide project funded by the [Department for Transport](https://www.gov.uk/government/organisations/department-for-transport) and transport charity [Sustrans](https://www.sustrans.org.uk/) to help relieve immediate pressures on the transport system and create long term change. The result is the [Rapid Cycleway Prioritisation Tool](https://www.cyipt.bike/rapid/). This is a free and open tool to help ensure that the government's [Emergency Active Travel Fund](https://www.gov.uk/government/publications/emergency-active-travel-fund-local-transport-authority-allocations) is spent where it is most needed, for maximum long term benefit.

## New routes

The [Rapid Cycleway Prioritisation Tool](https://www.cyipt.bike/rapid/) provides an interactive map for every transport authority in England. The main result is a list of road sections that have both high demand and sufficient space for cycling.

These are represented as blue lines in the map of Leeds below, many of which are included in the council's [new cycleway plans](https://news.leeds.gov.uk/news/leeds-city-council-announces-emergency-walking-and-cycling-plans-in-response-to-covid-19). Additional layers in the map include existing off road cycleways (shown in green), which in many cities are disjointed or of variable quality, and, in purple, a vision of what a joined-up cycle network could look like.

<image id="340974" align="centre" source="University of Leeds" caption="The map output of the Rapid Cycleway Prioritisation Tool for West Yorkshire, zoomed in on Leeds" />

This new tool builds on [our previous work](https://www.jtlu.org/index.php/jtlu/article/view/862), carried out with other universities, to help authorities identify and develop [strategic cycle networks](https://npct.github.io/pct-shiny/regions_www/www/static/03d_other_reports/2019-use-of-pct-report.pdf). We created a national dataset of roads, including estimates of the number of lanes in either direction and road width.

This can help identify where roads have space that could be reallocated to widen pavements or to rapidly introduce new cycleways. One of the first of has been installed outside a hospital in [Leicester](https://www.leicestermercury.co.uk/news/leicester-news/light-lockdown-traffic-leicester-allows-4083625). Community engagement through projects such as [Widen My Path](https://www.widenmypath.com/), which provides a forum for comments on where more space for walking and cycling is most needed, is another vital part of this process.

<tc-read-more id="135248"></tc-read-more>

The tool's focus on road space reallocation is due to the importance of speed and capacity. Light segregation measures such as plastic bollards or [wands](https://www.bbc.co.uk/news/uk-england-leeds-52577554) can be implemented much faster than constructing new cycleways from scratch. Wide, direct and continuous cycleways, of the type created by road space reallocation on wide roads, are needed for capacity and to ensure physical distancing guidelines can be followed. Finally, road space reallocation has environmental benefits, representing '[zero carbon infrastructure](https://www.ice.org.uk/news-and-insight/the-infrastructure-blog/october-2019/net-zero-needs-zero-carbon-infrastructure)'.

An important [finding](https://osf.io/7wjb6) is that many cities have wide roads with spare space, as shown the maps of six major cities below. Notably, none of these yet have a joined-up cycle network. An example of the type of road highlighted by our tool is [Kirkstall Road](https://www.openstreetmap.org/way/7991370#map=15/53.8049/-1.5718) in Leeds, which has high cycling potential and sufficient width for new cycleways. Kirkstall road is already part of plans by [Leeds City Council](https://news.leeds.gov.uk/news/new-measures-to-support-social-distancing-for-people-walking-and-cycling-in-leeds) to become a [trial cycleway](https://www.transportxtra.com/publications/local-transport-today/news/65382/leeds-to-pilot-enhanced-cycling-safety-u).

Our methods, based on open data and code, could be used in cities worldwide. Given the global nature of the challenges we face, open and collaborative research is vital. The potential for international application can be seen in [research we carried out](https://atfutures.github.io/upthat/articles/adaptation.html) for the World Health Organisation on possible cycling uptake in cities in low income countries.

<image id="341363" align="centre" source="Robin Lovelace" caption="Maps showing existing, disjointed cycleway networks (green), potential cycleway routes on wide roads according to the Rapid Cycleway Prioritisation Tool (blue) and cohesive networks (purple) in 6 major cities." zoomable="true" />

## Planning for the future

The UK government has announced £2 billion of investment in measures to promote walking and cycling in England over the next five years. £250 million has been allocated for emergency interventions [to make cycling and walking safer](https://www.gov.uk/government/speeches/transport-secretarys-statement-on-coronavirus-covid-19-9-may-2020).

Similar commitments are being made in [Scotland](https://www.transport.gov.scot/news/10-million-to-support-pop-up-active-travel-infrastructure/) and [Wales](https://gov.wales/councils-challenged-change-wales-transport-environment-during-coronavirus-lockdown). Local authorities urgently need to decide how this funding should be spent.

If action is prioritised in places where there is a long-term need for cycle improvements, there is a greater chance that these developments can become permanent. In Paris, COVID-19 related measures aim to contribute towards Mayor Anne Hidalgo's long-term [Plan Velo](https://www.curbed.com/2020/1/15/21065343/bike-paris-cycling-anne-hidalgo).

@@

New cycling infrastructure is more likely to be effective when it is developed based on analysis of the best available data, in combination with vital local knowledge. City planners, politicians and citizens need to act to ensure that transport interventions made during the crisis are of maximum benefit now and in the post-pandemic world.

@@

<!-- ![](https://raw.githubusercontent.com/cyipt/popupCycleways/master/README_files/figure-gfm/manchester-1.png) -->
